# Approaches to learning: data analysis


## Aim of the project and data

The project explores the impact of various factors on learning outcomes. To achieve this aim a survey was conducted among statistics course students in 2014. The researchers collected information about gender and age of the participants. The survey asked the participants to indicate their attitude towards statistics, and fill in the questionnaire describing their learning habits. The answers were divided into three categories, describing general ways of learning. We have called them deep, surface and strategic learning. Finally, the final mark of the course, given in points, was included in the data. After initial processing of the data, we have produced a data set, consisting of 166 observations over 7 variables.

```{r}
learning2014 <- read.table("Z:\\Documents\\RProjects\\IODS-project\\data\\learning2014.csv", header = TRUE, sep = ",", row.names = 1)

dim(learning2014)
str(learning2014)

```
## Summary statistics

As you can see from the graph below, female participants comprise the majority of the course. They tended to be only slightly younger than their male counterparts, and slightly more successful in the final grading. As for leaning habits, both male and female students show similar affiliation with deep and strategic learning, whereas surface learning is markedly more preset within the male population. Another notable difference between genders can be seen in attitude, with male students expressing higher regard for statistics than female students.

In our further analysis we plan to investigate the impact on learning outcomes, that is, we are treating our *points* variable as dependent variable. Our analysis of the correlation between different variables indicates that *attitude* the strongest and most statistically significant relationship to *points*. Less significant, but still of interest, is potential positive association of the course mark with strategic learning and negative one with surface learning. These relationships make common sense: positive attitude towards the study subject and strategic approach to learning result in higher marks, while surface learning allows very limited understanding of the subject and will probably lead to a lower grade. In the next secrion we will inroduce some tests to these assumptions.

```{r}
#changing the order of rows
library(dplyr)
learning2014 <- select(learning2014, one_of(c("gender","age","attitude","deep", "stra", "surf", "points")))

library(ggplot2)
library(GGally)

ggpairs(learning2014, mapping = aes(col=gender, alpha = 0.3), upper = list(continuous = wrap("cor", size=3)), lower = list(continuous = wrap("points", size = 0.3), combo = wrap("facethist", bins = 20)))

```
Insert para 4 fitting model

```{r}
model_learning2014 <- lm(points~ attitude + stra + surf, data = learning2014)

summary(model_learning2014)
```

Insert para 5 fitting the model further

```{r}
model2_learning2014 <- lm(points~ attitude, data = learning2014)

summary(model2_learning2014)
```

Insert para 6

```{r}
par(mfrow = c(2,2))
plot(model2_learning2014, which=c(1,2,5))

```