# Insert chapter 2 title here

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

Insert para 1 describing describing dimensions and structure

```{r}
learning2014 <- read.table("Z:\\Documents\\RProjects\\IODS-project\\data\\learning2014.csv", header = TRUE, sep = ",", row.names = 1)

dim(learning2014)
str(learning2014)

```

Insert para 2 describing graphical overview

```{r}
#changing the order of rows
library(dplyr)
learning2014 <- select(learning2014, one_of(c("gender","age","attitude","deep", "stra", "surf", "points")))

library(ggplot2)
library(GGally)
ggpairs(learning2014, mapping = aes(col=gender, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))

```
Insert para 4 fitting model

```{r}
model_learning2014 <- lm(points~ attitude + stra + surf, data = learning2014)

summary(model_learning2014)
```

Insert para 5 fitting the model further

```{r}
model2_learning2014 <- lm(points~ attitude, data = learning2014)

summary(model2_learning2014)
```

Insert para 6

```{r}
par(mfrow = c(2,2))
plot(model2_learning2014, which=c(1,2,5))

```