# Applying clustering methods to Boston data set


## Overview of the data

For this task I am using *Housing Values in Suburbs of Boston*, which is included into MASS package. The data set consists of 506 observations and 14 variables. See more information about the data [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html).

Summary of the variables reveals a few striking characteristics of Bostos suburbs. Per capita crime rate `crim` shows that majority of suburbs have relatively low crime rate with a few outliers. The measure  `black` reveals that majority of suburbs has similar proportions of black population but there is a group of suburbs with very amount of black residents.

Correlation plots reveals that crime rate is strongly associated with accessibility of radial highways, `rad`,  with full value property tax rate `tax`. These variables also highly dependent from each other. Relatively strong negative relationships are visible with distances to employment centers (`dis`), proportion of black residents (`black`), and median value of owner-occupied homes (`medv`). Suburbs with high grime rates also seem to located near industrial centers (`indus`) and,prbably as a result, be more polluted (`nox`).

```{r}
library(MASS)
library(dplyr)
library(tidyr)
library(corrplot)

data("Boston")

# structure and dimensions of the data
glimpse(Boston)

# summaries of variables
summary(Boston)

# graphical overview

cor_matrix<-cor(Boston) %>% round(digits = 2)

## measuring statistical significance of correlation coefficients
testRes <- cor.mtest(cor_matrix, conf.level = 0.95)

## plotting with significant level stars
corrplot(cor_matrix, p.mat = testRes$p, method="square", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6, sig.level = c(0.001, 0.01, 0.05), pch.cex = 0.9, insig = 'label_sig', pch.col = 'black')

```

## Preparing the data set for future analysis

### Scaling

For the purposes of lda analysis, I am first scaling the data set. As you can see, now the means of all variables are zero, so it is easy now to compare their "intensity".

```{r}
# centering and standardizing variables
boston_scaled <- scale(Boston)

boston_scaled <- as.data.frame(boston_scaled)

summary(boston_scaled)

```

### Creating crime rate target variable

In order conduct lda analysis with crime rate as a target variable, I will need to transform it to a categorical data. I am going to create categories based on quantiles, but as the summary of this variable has indicated earlier, I am expecting to see a significant break in the third quantile.

```{r}
# creating a quantile vector
bins <- quantile(boston_scaled$crim)

# creating a categorical variable 'crime'
values <- c("low", "med_low", "med_high", "high")
crime <- cut(boston_scaled$crim, breaks = bins, label = values, include.lowest = TRUE)

# removing original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# adding the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

summary(boston_scaled$crime)

```

### Dividing to train and test sets

Finally, I am dividing the set, allocating 80% as trai data, and the rest as test data.

```{r}
# number of rows in the Boston data set 
n <- nrow(boston_scaled)

# selecting randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# train set
train <- boston_scaled[ind,]

# test set 
test <- boston_scaled[-ind,]

# saving the correct classes from test data
correct_classes <- test$crime

# removing the crime variable from test data
test <- dplyr::select(test, -crime)
```

## LDA analysis

```{r}
lda.fit <- lda(crime ~ ., data = train)

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plotting the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1.5)
```

